---
title: "Olympic Medals"
author: "Gabriela Diaz, Felicia Peterson, & Gabby Valle"
date: "4/12/2022"
output: html_document
---

```{r}
#knitr::opts_chunk$set(echo = TRUE, error=TRUE, message=FALSE, warning=FALSE)
```



```{r libraries}
library(tidyverse)     # for data cleaning and plotting
library(lubridate)     # for date manipulation
library(openintro)     # for the abbr2state() function
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
library(gplots)        # for col2hex() function
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(leaflet)       # for highly customizable mapping
library(ggthemes)      
library(plotly)        # for the ggplotly() - basic interactivity
library(gganimate)     # for adding animation layers to ggplots
library(gifski)        # for creating the gif (don't need to load this library every time,but need it installed)
library(transformr)    # for "tweening" (gganimate)
library(shiny)         # for creating interactive apps
library(patchwork)     # for nicely combining ggplot2 graphs
library(gt)            # for creating nice tables
library(countrycode)
theme_set(theme_minimal())

```

## Original Data

```{r}
olympics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-07-27/olympics.csv')

world <- map_data("world")

medal_colors <- c(
  Gold = "#d5a500",
  Silver = "#b7b7b7",
  Bronze = "#a17419")

```

## Filtered Data Sets

```{r}
summer_olympics <- olympics %>% 
  filter(season == "Summer") %>% 
   mutate(countryName = countrycode(noc, "genc3c", "country.name")) %>% 
   mutate(countryName = ifelse(countryName == "USA", "United States", countryName)) %>% 
   mutate(countryName = ifelse(noc == "ALG", "Algeria", countryName)) %>% 
   mutate(countryName = ifelse(noc == "BAH", "Bahamas", countryName)) %>%
   mutate(countryName = ifelse(noc == "BUL", "Bulgaria", countryName)) %>%
   mutate(countryName = ifelse(noc == "CHI", "Chile", countryName)) %>%
   mutate(countryName = ifelse(noc == "CRC", "Costa Rica", countryName)) %>%
   mutate(countryName = ifelse(noc == "CRO", "Croatia", countryName)) %>%
   mutate(countryName = ifelse(noc == "DEN", "Denmark", countryName)) %>%
   mutate(countryName = ifelse(noc == "FIJ", "Fiji", countryName)) %>%
   mutate(countryName = ifelse(noc == "GER", "Germany", countryName)) %>%
   mutate(countryName = ifelse(noc == "GRE", "Greece", countryName)) %>%
   mutate(countryName = ifelse(noc == "GRN", "Grenada", countryName)) %>%
   mutate(countryName = ifelse(noc == "HAI", "Haiti", countryName)) %>%
   mutate(countryName = ifelse(noc == "INA", "Indonesia", countryName)) %>%
   mutate(countryName = ifelse(noc == "IRI", "Iran", countryName)) %>%
   mutate(countryName = ifelse(noc == "LAT", "Latvia", countryName)) %>%
   mutate(countryName = ifelse(noc == "MGL", "Mongolia", countryName)) %>%
   mutate(countryName = ifelse(noc == "NED", "Netherlands", countryName)) %>%
   mutate(countryName = ifelse(noc == "NGR", "Nigeria", countryName)) %>%
   mutate(countryName = ifelse(noc == "POR", "Portugal", countryName)) %>%
   mutate(countryName = ifelse(noc == "PUR", "Puerto Rico", countryName)) %>%
   mutate(countryName = ifelse(noc == "RSA", "South Africa", countryName)) %>%
   mutate(countryName = ifelse(noc == "SLO", "Slovenia", countryName)) %>%
   mutate(countryName = ifelse(noc == "SUI", "Switzerland", countryName)) %>%
   mutate(countryName = ifelse(noc == "UAE", "United Arab Emirates", countryName)) %>%
   mutate(countryName = ifelse(noc == "URU", "Uruguay", countryName)) %>%
   mutate(countryName = ifelse(noc == "VIE", "Vietnam", countryName)) %>%
   mutate(countryName = ifelse(noc == "ZIM", "Zimbabwe", countryName)) %>% 
drop_na(countryName)

# Footnote or mention in part of the project the choice to leave out territories not fully recognized, such as Kosovo and Chinese Taipei, and countries that no longer exist, including the Soviet Union and Yugoslavia. 


summer_gold <- summer_olympics %>% 
  drop_na(medal) %>% 
  filter(medal == "Gold") 

summer_medals <- summer_olympics %>% 
  drop_na(medal)

medal_prop <- summer_olympics %>% 
  count(countryName, medal) %>% 
  group_by(countryName) %>% 
  filter(any(!is.na(medal))) %>% 
  mutate(medal = ifelse(is.na(medal), "No Medal", medal)) %>% 
  pivot_wider(
    names_from = medal,
    values_from = n,
    values_fill = 0) %>% 
  mutate(
    total_medals = Gold + Silver + Bronze,
    total_events = total_medals + `No Medal`,
    p_medal = total_medals / total_events) %>% 
  arrange(desc(total_medals))




```

Female-only, medal-winning data:

```{r}

golden_girls <- summer_gold %>% 
  filter(sex == "F") %>% 
  drop_na(height, weight) 


```
Instructions that might be useful to remember:

* **NEW!!** With animated graphs, add `eval=FALSE` to the code chunk that creates the animation and saves it using `anim_save()`. Add another code chunk to reread the gif back into the file. See the [tutorial](https://animation-and-interactivity-in-r.netlify.app/) for help. 

* When you are finished with ALL the exercises, uncomment the options at the top so your document looks nicer. Don't do it before then, or else you might miss some important warnings and messages.


##Taking data out "on a walk" (aka brainstorming)

```{r}

summer_gold %>%
  group_by(countryName) %>% 
  count(medal) %>% 
  arrange(desc(n)) #results in a medal count in descending order

```


##Felicia's Graphs

1. Overall Medals won

```{r, fig.height=5, fig.alt= "words words words Reflects women not being allowed in competitions for a good portion of all Olympic Games."}
summer_olympics %>%
  group_by(sex, sport) %>% 
  count(medal) %>% 
  filter(!(sport %in% c("Aeronautics", "Art Competitions", "Jeu De Paume", "Roque", "Alpinism", "Basque Pelota", "Croquet", "Racquets", "Figure Skating", "Ice Hockey", "Cricket", "Polo", "Trampolining", "Tug-Of-War", "Motorboating"))) %>% 
  ggplot(aes(x = n, y = reorder(sport, n), fill = sex)) + 
  geom_col() +
  labs(y = "",
        title = "Overall medals won by men and women Olympians")


```




2. Women's heights and weights

```{r}
#weight in kg
# Athletics = track and field
# Hockey = Field Hockey
#Leaving out figure skating because  1920s Olympics, one gold winner

golden_girls <- summer_olympics %>% 
  filter(sex == "F") %>% 
  drop_na(height, weight) 
```


```{r}
golden_girls %>% 
  ggplot(aes(x = weight, y = height, color = sport)) +
  geom_point() 
```





```{r, fig.height=7, fig.alt= "Here is the description yippeeeee!" }
equip_gg <- golden_girls %>% 
  filter(sport == c("Badminton", "Cycling", "Equestrianism", "Golf",  "Hockey", "Shooting", "Softball", "Table Tennis", "Tennis", "Rhythmic Gymnastics", "Weightlifting"))
equip_gg

armor_gg <- golden_girls %>% 
  filter(sport == c("Boxing", "Judo", "Fencing", "Football", "Taekwondo", "Rugby Sevens"))
armor_gg

hydro_gg <- golden_girls %>% 
  filter(sport == c("Canoeing", "Diving", "Water Polo", "Rowing", "Swimming", "Synchronized Swimming", "Sailing"))
hydro_gg

solo_gg <- golden_girls %>%
  filter(sport == c("Athletics", "Basketball", "Beach Volleyball", "Gymnastics", "Handball", "Trampolining", "Triathalon", "Volleyball", "Wrestling" ))
solo_gg

fig1 + fig2 + fig3 + fig4 + 
  plot_annotation(title = "Olympic Medal Winning Women",
                  subtitle = "Trends in height and weight for female athletes depending on sport")
```



3. World map showing the proportion of Olympic medal wins to participation in events

```{r}

medal_prop %>% 
  ggplot() +
  geom_map(map = world, 
           aes(map_id = countryName,
               fill = p_medal)) +
  scale_fill_gradient(low=cm.colors(20)[1], high=cm.colors(20)[20])+
  expand_limits(x = world$long, y = world$lat) +
  theme_map()

 
```




