---
title: "Olympic Medals"
author: "Gabriela Diaz, Felicia Peterson, & Gabby Valle"
date: "4/12/2022"
output: html_document
---

```{r}
#knitr::opts_chunk$set(echo = TRUE, error=TRUE, message=FALSE, warning=FALSE)
```



```{r libraries}
library(tidyverse)     # for data cleaning and plotting
library(lubridate)     # for date manipulation
library(openintro)     # for the abbr2state() function
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
library(gplots)        # for col2hex() function
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(leaflet)       # for highly customizable mapping
library(ggthemes)      
library(plotly)        # for the ggplotly() - basic interactivity
library(gganimate)     # for adding animation layers to ggplots
library(gifski)        # for creating the gif (don't need to load this library every time,but need it installed)
library(transformr)    # for "tweening" (gganimate)
library(shiny)         # for creating interactive apps
library(patchwork)     # for nicely combining ggplot2 graphs
library(gt)            # for creating nice tables
library(countrycode)
theme_set(theme_minimal())

```

## Original Data

```{r}
olympics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-07-27/olympics.csv')


```

## Filtered Data Sets

```{r}
summer_gold <- olympics %>% 
  filter(season == "Summer") %>% 
  drop_na(medal) %>% 
  filter(medal == "Gold") 

summer_gold
  
 
  group_by(noc) %>% 
  count(medal) %>% 
  arrange(desc(n))



```


Instructions that might be useful to remember:

* **NEW!!** With animated graphs, add `eval=FALSE` to the code chunk that creates the animation and saves it using `anim_save()`. Add another code chunk to reread the gif back into the file. See the [tutorial](https://animation-and-interactivity-in-r.netlify.app/) for help. 

* When you are finished with ALL the exercises, uncomment the options at the top so your document looks nicer. Don't do it before then, or else you might miss some important warnings and messages.
